Technique for \earlywarning\ detection of compact binary coalescence and its implication for multi-messenger astronomy

### 2011-06-03 method.tex
Minor edits in section 2
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@887 ccdd05d5-56c1-461a-bf98-85f6876523b4

In the remainder of this section we explore a method for reducing the
computational cost of a time domain search for compact binary coalescence.  We
will show that it is possible to have a theoretically zero latency algorithm
that competes with the computational cost of the standard frequency domain
matched filtering, which requires a signficant latency in order to be
computationally cheap. Our method, \lloid\ involves two transformations of the
template waveforms that produce a set of orthogonal filters with far fewer
sample points than the original template waveforms.  The reduction in the
number of sample points in all of the filters required to search the entire
parameter space is dramatic.

The first transformation is to chop the time domain templates up into time
slices.  Since each template piece is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim 10$ samples on
average and also allows the filters to be evaluated at about a $\sim 10$ times
lower rate on average.  This amounts to more than a factor of 100 in the
floating point operations per second required.  However, the resulting filters
are still not orthogonal across the mass parameter space, and are in fact
highly redundant.  We use the singular value decomposition to produce an
orthogonal filter set from the time sliced templates~\cite{Cannon:2010p10398}.
We find that this reduces the number of sample points that need to be filtered
by another factor of $\sim 100$.  The combined methods reduce the number of
floating point operations to the level where they are competitive with the
conventional matched filter approach.  In the remainder of this section we
describe the \lloid\ algorithm in detail and provide some basic computational
cost scaling.

@@

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{-t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency $f_\textrm{hi}$.
This is often chosen to correspond to the \ISCO. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t^0)$ and is said to have  a
duration of $t^0 - t_\mathrm{low}$. It is critically sampled at a
rate of $2 f_\mathrm{hi}$.

The monotonic time-frequency relationship of \eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the filter integral at a lower sample
rate.  Specifically we consider here time slice boundaries with the next
highest power-of-two sample rate that critically samples the time sliced
template.  The time slices consist of the $S$ intervals $\left(t^S,
t^{S-1}\right],\, \dots,\, \left(t^2, t^1\right],\, \left(t^1, t^0\right]$ sampled at frequencies
$f^\mathrm{S-1},\, \dots,\, f^1,\, f^0$ where $f^0 \geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ are the number of sample points in the
$s^{\mathrm{th}}$ filter.}
\end{table}

>>>>

In the remainder of this section we explore a method for reducing the
computational cost of a time domain search for compact binary coalescence.  We
will give a truly zero latency algorithm that competes in terms of floating point
operations per second with the conventional frequency domain method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of orthogonal
filters with far fewer coefficients than the original templates.  The reduction
in the number of sample points in all of the filters required to search the entire
parameter space is dramatic.

The first transformation is to chop the time domain templates up into time
slices.  Since each template piece is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim 10$ samples on
%
\editorial{Where do these numbers come from?}
%
average and also allows the filters to be evaluated at about a $\sim 10$ times
lower rate on average.  This amounts to more than a factor of 100 in the
floating point operations per second required.  However, the resulting filters
are still not orthogonal across the mass parameter space, and are in fact
highly redundant.  We use the singular value decomposition to produce an
orthogonal filter set from the time sliced templates~\cite{Cannon:2010p10398}.
We find that this reduces the number of sample points that need to be filtered
by another factor of $\sim 100$.  The combined methods reduce the number of
floating point operations to the level where they are competitive with the
conventional matched filter approach.  In the remainder of this section we
describe the \lloid\ algorithm in detail and provide some basic computational
cost scaling.

@@

For concreteness and simplicity, we will consider an inspiral waveform in the
quadrupole approximation, for which the time-frequency relation is
%
\begin{equation} \label{eq:fgw}
%
f = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \umu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary~\cite{findchirppaper, kidder1992}.  Usually the
template is truncated at some prescribed time $t^0$, or equivalently frequency $f_\textrm{hi}$.
This is often chosen to correspond to the \ISCO. An inspiral signal
will enter the detection band at a low frequency, $f = f_\mathrm{low}$,
corresponding to a time $t_\mathrm{low}$.  The template is assumed to be zero
outside the interval $[t_\mathrm{low}, t^0)$ and is said to have  a
duration of $t^0 - t_\mathrm{low}$. It is critically sampled at a
rate of $2 f_\mathrm{hi}$.

The monotonic time-frequency relationship of equation~\eqref{eq:fgw} allows us to choose
time-slice boundaries that require substantially less bandwidth at early times
in the inspiral.  Our goal is to reduce the filtering cost of a large fraction
of the waveform by computing part of the convolution at a lower sample
rate.  Specifically we consider here time slice boundaries with the smallest
power-of-two sample rate that subcritically samples the time sliced
template.  The time slices consist of the $S$ intervals $\left(t^S,
t^{S-1}\right],\, \dots,\, \left(t^2, t^1\right],\, \left(t^1, t^0\right]$ sampled at frequencies
$f^\mathrm{S-1},\, \dots,\, f^1,\, f^0$ where $f^0 \geqslant 2 f_\mathrm{hi}$ and $f^{S-1} \geqslant 2 f_\mathrm{low}$.  The time sliced templates may be downsampled without aliasing, so we define them as
%
\begin{equation}
\label{eq:time-sliced-templates}
h_{i}^{s}[k] \equiv
	\cases{
		h_{i}\!\left[k\frac{f}{f^s}\right] & if $t^s \leqslant k/f^s < t^{s+1}$ \\
		0 & otherwise.
	}
\end{equation}
%
An example time-slice design satisfying these constraints for a $1.4 - 1.4 \, M_{\odot}$ binary is shown in table~\ref{table:time_slices}.
%
\begin{table}[h!]
\begin{minipage}[c]{0.52\textwidth}
\centering
\vspace{0.8cm}
\includegraphics{time_slices.pdf}
\end{minipage}
\begin{minipage}[c]{0.3\textwidth}
\centering
\input{time_slices.tex}
\end{minipage}
\caption{\label{table:time_slices} Example of nearly critically sampled,
power-of-two time slices for a $1.4 - 1.4 \, M_{\odot}$ template extending from
$f_\mathrm{low} = 10 \, \mathrm{Hz}$ to $f_\mathrm{ISCO} = 1571\, \mathrm{Hz}$
with a time frequency structure given by ($\ref{eq:fgw})$. $f^s$ is the sample
rate of the time slice, $(t^{s+1}, t^s]$ are the boundaries in seconds
preceeding coalescence and \slicessamps\ is the number of samples in the
$s^{\mathrm{th}}$ filter.}
\end{table}
